%\input{commands}
\chapter{Decomposing primordial shapes}
\section{Pulling out the $k$-dependance}
    \subsection{Set up formalism.}
    \subsection{Forced to decompose on the cube.}
    \subsection{Defining the notation for the two main time integrals.}
\section{Inner product choice}
    \subsection{Easier to interpret, more stringent.}
    \subsection{Notes on observable convergence being the deciding factor.}
\section{Template testing}
    \subsection{Need to understand expected convergence before validation on numerical results.}
\section{Decomposing shapes on the cube vs tetrapyd}
    \subsection{Large non-physical contributions.}
    \subsection{FIGURE: DBI on cube vs tetra.}
\section{Setting up a basis: augmentation}
    \subsection{Using modified GS.}
    \subsection{Legendre and Fourier as building blocks}
\section{Basis choice matters!}
    \subsection{Why the basic basis expansion is so bad.}
    \subsection{My new basis sets, and their dramatic improvement.}
    \subsection{Nice table with descriptions and some single-number comparison on examples.}
    \subsection{FIGURES: Recon for Malda, DBI, scale-inv, with P0, P1, F0, F1 vs $P_{max}$.}
    \subsection{FIGURES: Recon for cos, cosDBI, with P0, P1, F0, F1 vs $P_{max}$.}
    \subsection{FIGURES: Recon for cos-log, cos-logDBI, with P0, P1, F0, F1 vs $P_{max}$.}
    \subsection{FIGURES: Recon for scaled-DBI, with P0, P1, P1ns, P01ns, $P_log$ vs $P_{max}$.}
\section{Not obvious, but $k_{ratio}$ matters!}
\section{The $P_{max}$-$k_{ratio}$ tradeoff (and DBI vs Equil).}
\section{Log basis for basis without PS info.}
\section{Loginv basis for resonant shapes.}
\section{Factor basis?}
\section{Conclusion}
    \subsection{Basic Legendres/Fourier are not sufficient.}
    \subsection{Need to pay attention to the cube.}
    \subsection{Including 1/k terms helps massively.}
    \subsection{The log basis performs excellently, without being tied to a specific $n_s$.}
