%\input{commands}
\chapter{Constraints TO DO}\label{chapter:constraints}
\section{Connecting to CMB-BEst}
    The goal here is to validate our pipeline by reproducing the $\planck$~result
    from~\cite{Planck_NG_2018}, equation (55):
    \begin{align}
        c_s^{DBI}&\ge0.079\qquad(95\%,~T~\text{only}).
    \end{align}
    We would not expect to reproduce this exactly due to
    \textcolor{red}{Lack of maps? Different approximations? Numerics?}
    Though we can estimate a reasonable disagreement by looking at
    the expected scatter between the Planck estimators, and by calculating
    \textcolor{red}{$\sigma$??}.
    We use equation~\eqref{eq:dbi_warp}, with 
\section{Compare convergence at primordial level to convergence at $f_{NL}$ level}
    We will compare primordial convergence to CMB convergence,
    by comparing the $\Pmax=29$ and $\Pmax=30$ results.
    This will show that the CMB convergence is slightly better, i.e.\ that
    the squeezed limit (which is where the primordial shape converges most slowly)
    is supppressed.
    For DBI, and also for DBI resonance?
\section{Validate this convergence by reproducing Planck constraint using DBI template decomp.}
    Words.
\section{Then, check how the scaling from the real numerical result affects this.}
    Here we will compare template decompositions to Primodal results.
    This will tell us how large the slow-roll corrections are to the
    final CMB result, but not anything about the primordial convergence.
%\section{EFT stuff, build off Enrico's recent work.}
%    Words
