%\input{commands}
\chapter{Introduction to bispectra}
\section{Inflation}
    \subsection{Perturbations}
    \subsection{$c_s$.}
    \subsection{Bunch-Davies initial conditions.}
    \subsection{MS equation, behaviour at various times.}
    \subsection{MS equation in $N$, $\tau_s$ - no numerics here, though, just ref forward.}
    \subsection{Define the power spectrum, discuss constraints.}
\section{The primordial bispectrum}
    \subsection{Isotropy etc.}
    \subsection{f_{NL} (various definitions).}
    \subsection{Link inflaton non-linearity to primordial NG.}
    \subsection{Link primordial NG to the CMB, possibly with some simple example.}
\section{Calculating the bispectrum}
    \subsection{Coming from part iii level, a derivation of the tree-level in-in formalism}
    \subsection{Calculate the interaction Hamiltonian}
    \subsection{Through the self-interactions, i.e. neglecting the metric perturbations.}
    \subsection{Discuss DBI, then $P(X, \phi)$.}
    \subsection{Review of the Maldacena calculation, with metric perturbations.}
    \subsection{Discuss field-redefinition being unnecessary, as per 1103.4126.}
    \subsection{Examples of shapes, using $K_{pq}$ notation (link to Enrico's sym polys?)}
    \subsubsection{Basic shapes, $f_{NL}$}
    \subsubsection{Maldacena, DBI}
    \subsubsection{$P(X, \phi)$, EFT}
    \subsubsection{Features, with explicit details of how resonance and features generate large NG.}
    \subsection{F&RP's work}
    \subsubsection{Summary of the achievements and limitations, how I went beyond them.}
    \subsection{Previous config-by-config codes}
    \subsubsection{Their limitations.}
    \subsubsection{Their usage in recent works.}
    \subsection{The squeezed limit consistency condition.}
\section{Bispectrum data analysis}
    \subsection{Simons, Planck, projections}
    \subsection{Statistical observables}
    \subsection{Estimators.}
    \subsection{Determining the weighting of a coin.}
    \subsection{Link to Wuhyun.}
    \subsection{Estimating the bispectrum, complexity.}
    \subsection{Review estimators, KSW, separability.}
    \subsection{Separable approximations to non-separable templates}
    \subsection{Equilateral to DBI.}
    \subsection{Approximations to approximations.}
    \subsection{Modal methods, constraints from Planck.}
    \subsection{Wuhyun's stuff as development of Modal methods (cite Wuhyun paper)}
