%%% ps2ascii thesis.pdf | wc -w
%%% bc <<< "scale=2; 5./60000."
%%% word_count=$(ps2ascii thesis.pdf | wc -w) && bc <<< "scale=2; ${word_count}/60000."
%%% https://www.cst.cam.ac.uk/local/phd/typography
%%% https://www.cst.cam.ac.uk/local/phd/writingup


%%%%%
%%
%% Sample document ``thesis.tex''
%%
%% Version: v0.2
%% Authors: Jean Martina, Rok Strnisa, Matej Urbas
%% Date: 30/07/2008
%%
%% Copyright (c) 2008-2011, Rok Strniša, Jean Martina, Matej Urbas
%% License: Simplified BSD License
%% License file: ./License
%% Original License URL: http://www.freebsd.org/copyright/freebsd-license.html
%%%%%

% Available documentclass options:
%
%   <all `report` document class options, e.g.: `a5paper`>
%   withindex   - enables the index. New index entries can be added through `\index{my entry}`
%   glossary    - enables the glossary.
%   techreport  - typesets the thesis in the technical report format.
%   firstyr     - formats the document as a first-year report.
%   times       - uses the `Times` font.
%   backrefs    - add back references in the Bibliography section
%
% For more info see `README.md`
\documentclass[withindex,glossary]{cam-thesis}

% Citations using numbers
%\usepackage[numbers]{natbib}
\usepackage[numbers,sort&compress]{natbib}
\usepackage{subfig}
\usepackage[dvipsnames]{xcolor}
\usepackage{booktabs}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Thesis meta-information
%%

%% The title of the thesis:
%\title{Probing Inflation with Precision Bispectra}
\title{Connecting Inflation to Observations Through the Bispectrum}

%% The full name of the author (e.g.: James Smith):
\author{Philip Clarke}

%% College affiliation:
\college{St.\ John's}

%% College shield [optional]:
\collegeshield{CollegeShields/StJohns}

%% Submission date [optional]:
% \submissiondate{November, 2042}

%% You can redefine the submission notice [optional]:
% \submissionnotice{A badass thesis submitted on time for the Degree of PhD}

%% Declaration date:
\date{July, 2021}

%% PDF meta-info:
\subjectline{Early-universe cosmology}
\keywords{inflation}



%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Abstract:
%%
\abstract{%
\textcolor{red}{This is just a placeholder outline, will be reduced to proper abstract size eventually.}
%    \textbf{Dissertation Summary.}
Calculating the full primordial bispectrum predicted by a model of inflation and
comparing it to what we see in the sky is very computationally intensive, necessitating layers
of approximations and limiting the models which can be constrained. The inherent separability
of the tree level in-in formalism provides a means by which to
obviate some of these difficulties.
To exploit this property, one can expand in separable basis functions.
The practicality of this method is then determined by the descriptive
power of the basis chosen, i.e.\ by the range of scenarios for which that basis provides a convergent representation
of the bispectrum. The central difficulty encountered in obtaining fast convergence is the effect of dominant non-physical
$k$-configurations. A secondary difficulty encountered is accurately including the early-time contributions
to the higher-order coefficients (which are necessary to capture feature effects, such as resonance non-Gaussianity).


In this thesis we develop this separable approach into a practical and efficient
numerical methodology which can be applied to a much wider and more complicated range of bispectrum
phenomenology than previous analyses. This is an important step forward towards observational pipelines which can directly
confront specific models of inflation.


The specific results involved can be categorised into two main lines of research.
The first line of research is in determining the feasibility of the method.
Our initial result here was identifying the contributions of the non-physical configurations
as a novel problem to this analysis, and identifying basis choice as the key method of overcoming this difficulty.
In this work we describe multiple basis sets and make quantitative comparisons
of their convergence on realistic and interesting models, including ones with features.
We find basis sets that can overcome the difficulty of the dominant non-physical $k$-configurations,
converging far more efficiently than the basic basis sets, for physically interesting models.


The second line of research we follow is in developing basis-independent methods that allow the fast and accurate calculation
of higher order coefficients of the basis expansion of the tree-level in-in formalism.
To this end we set up the separable formalism and describe the methods we use to overcome the difficulties encountered.
These difficulties include accurately including the highly oscillatory early-time contributions.
We present a careful and comprehensive validation of our methods,
showing that we can overcome the convergence problems and capture the oscillatory contributions efficiently.
At the primordial level we do this by validating our results using three distinct tests.
We validate on established templates with non-trivial
phenomenology; we use the squeezed-limit consistency condition; and we compare our
full bispectrum results to previous codes through point-tests.


At the level of the CMB, we validate the convergence of our basis by ensuring we can reproduce the
$\planck$ constraints on the DBI sound speed, and we translate this constraint into a constraint on
a fundamental parameter in the context of a physically realistic scan.


Chapter~\ref{chapter:intro_general} is a general introduction to cosmology, and
chapter~\ref{chapter:intro_bispectra} is an introduction to the bispectrum as an observable.


Chapter~\ref{chapter:decomp} describes the first line of research mentioned above---we
discuss why we need to take the non-physical configurations into account, how we build our
basis sets, and present quantitative comparisons of the convergence of each basis set to
relevant examples of bispectrum shapes.


In chapter~\ref{chapter:methods} we detail the second line of research mentioned above---we
discuss the details of recasting the in-in formalism in an explicitly separable form,
making explicit each step of the calculation.
We discuss how to efficiently deal with the early time contributions to the integrals,
and other numerical issues that require considerable care and attention.
We also present validation tests on a very broad range of types of non-Gaussianity.


In chapter~\ref{chapter:constraints} we present constraints (obtained in collaboration with Wuhyun Sohn)
on the sound speed of the DBI model, which validate our pipeline against $\planck$.
This chapter will describe the parametrisation of the scan, and the physical
interpretation of the results.


In chapter~\ref{chapter:conclusion} we discuss possible avenues of future work, and present our conclusions.


\textcolor{red}{Put this somewhere:}
The standard cosmological history, the $\Lambda CDM$ model, explains the structure we see in the distribution of matter (large-scale structure, LSS) and radiation (the cosmic microwave background, \textit{CMB}) in our universe. However, this model still requires initial conditions. These initial conditions – adiabatic perturbations on a uniform background, close to Gaussian, with a power spectrum that is nearly (but not exactly) scale-invariant – are provided by an epoch of “inflation”, an epoch before the Big Bang of the $\Lambda CDM$ model. 

During inflation, the universe is supposed to have expanded rapidly, stretching quantum fluctuations from small scales to large. This seeded the formation of structure and encoded correlations that we would eventually measure today.
Much information on fundamental physics has been gleaned from the two-point correlators of observations. For higher order correlators (which probe the non-linearities of interactions in the early universe) calculating the predictions of a model of inflation and comparing them to what we see in the sky is much more computationally intensive. Previous work, including that done by the Planck collaboration~\cite{Planck_NG_2018}, has managed to constrain some models of inflation through their three-point correlations (or rather through their proxy, the bispectrum).
The computational complexity was dealt with by making layers of approximations and the space
of models which could be constrained was limited.
One notable simplifying assumption is that of “separability”, which makes the process of \textit{CMB} estimation much more tractable~\cite{Komatsu_2005}.


The formalism used to calculate inflationary non-Gaussianities within an inflation scenario is
known as the in-in formalism~\cite{weinberg_in_in}\ldots

}



%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Acknowledgements:
%%
\acknowledgements{%
    \textcolor{red}{Actually write this!}
  My acknowledgements\ldots\\
  We make use of Lyth and Liddle,\\
  https://nms.kcl.ac.uk/eugene.lim/AdvCos/lecture2.pdf,\\
  http://cosmology.amsterdam/education/cosmology/\\
  and~\cite{Baumann_tasi}\\
  throughout.
  I am very grateful to \textcolor{red}{lots of people} for \textcolor{red}{reasons}.
  St.\ John's College, the Cambridge Trust European Scholarship, STFC (and the taxpayers),
  the Cambridge Philosophical Society, the Robert Gardiner Scholarship.
}



%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Glossary [optional]:
%%
\newglossaryentry{HOL}{
    name=HOL,
    description={Higher-order logic}
}

%\input{commands}
\usepackage{commands}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Contents:
%%
\begin{document}



%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Title page, abstract, declaration etc.:
%% -    the title page (is automatically omitted in the technical report mode).
\frontmatter{}
%\thispagestyle{empty}
\listoffigures
\listoftables
%\newpage


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Chapter 1:
%%
\include{intro_chapter}
\include{intro2_chapter}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Chapter 2:
%%
\include{basis_chapter}
\include{primodal_chapter}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Chapter 3:
%%
%%\include{primordial_results_chapter}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Chapter 4:
%%
\include{constraints_chapter}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Chapter 5:
%%
\include{future_chapter}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% References:
%%
% If you include some work not referenced in the main text (e.g. using \nocite{}), consider changing "References" to "Bibliography".
%

% \renewcommand to change default "Bibliography" to "References"
\renewcommand{\bibname}{References}
\cleardoublepage
\phantomsection
\addcontentsline{toc}{chapter}{References}
%\nocite{*}
\bibliographystyle{JHEP}
\bibliography{thesis}



%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Appendix:
%%

\appendix

\chapter{Appendices}
\section{Efficient numerics}
Efficient evaluation and numerical usage of the coefficients (vander, etc).
Numeric choices that aren't specific to mode evolution, etc.
Gauss-Legendre integration, fixed weights.
\section{Tools used}
Numpy, Scipy, Jupyter Notebooks, QUADPTS...?
Cite~\cite{scipy, numpy}
\section{Transport method for the Modal coeffs}\label{appendix_modal_transport}
Infinite hierarchy of coupled equations.
Choose a basis to make it nice?
Real vs Imag...?
Sounds cool, not worth pursuing.
One could imagine applying the same philosophy to our method.
Certainly, at first sight this seems more natural, that if the core
quantities in our method are the coefficients in some basis expansion,
why not evolve them directly? Why take the apparently circuitous route
of evolving the $\zeta_k(\tau)$, and decomposing them at every timestep?
The answer is that the ``equations of motion'' for the coefficients of the expansion
obtained by substituting the mode expansion of $\zeta_k(\tau)$ into~\eqref{modefneqn_zeta_N}
are coupled in an infinite hierarchy.



%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Index:
%%
\printthesisindex

\end{document}
